\documentclass{article}

\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{hyperref}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\newcommand{\Z}{\mathbb{Z}}

\title{IMO 2016 N5}
\author{}
\date{}

\begin{document}

\maketitle



\subsection*{Problem}

Fix some $k, a \in \Z$ with $k \geq 0$ and $a > 0$.
A pair $(x, y) \in \Z^2$ is called \emph{nice} if $(k + 1) y^2 - k x^2 = a$.
Suppose that a nice pair exists.

\begin{enumerate}
\item   Prove that for any integer $N$, there exists a nice pair $(x, y)$ with $x \geq N$.
\item   Prove that there exists a nice pair $(x, y)$ with $x^2 \leq a$.
\end{enumerate}



\subsection*{Solution}

Official solution: \url{http://www.imo-official.org/problems/IMO2016SL.pdf}

We follow Solution 2 of the official solution.
The main idea is the following fact.
For any nice pair $(x, y)$, one can check that $(x', y') = ((2k + 1) x + (2k + 2) y, 2k x + (2k + 1) y)$ is also nice.
Indeed, by bashing,
\[ (k + 1) (2k x + (2k + 1) y)^2 - k ((2k + 1) x + (2k + 2) y)^2 = (k + 1) y^2 - k x^2. \]

\begin{enumerate}

\item
It suffices to show that, for any nice pair $(x, y)$, there exists a nice pair $(x', y')$ with $x' > x$.
Take $x' = (2k + 1) |x| + (2k + 2) |y|$ and $y' = 2k |x| + (2k + 1) |y|$, which works by the above fact.
It is easy to see that $x' > |x| \geq x$; note that $|y| > 0$ is forced.
This proves the first part.

\item
It suffices to show that if $x^2 > a$, then we can find another nice pair $(x', y')$ with $|x'| < |x|$.
This time, use $x' = (2k + 1) |x| - (2k + 2) |y|$ and $y' = 2k |x| - (2k + 1) |y|$.
Again, the above fact means that $(x', y')$ is nice.
Thus it remains to prove that
\[ |x'| < |x| \iff 2k |x| - (2k + 2) |y| < 0 < (2k + 2) |x| - (2k + 2) |y|. \]
For the latter, $x^2 > a$ and $(k + 1) y^2 - k x^2 = a$ implies $x^2 > y^2$ and thus $|x| > |y|$.
The former is equivalent to $k|x| < (k + 1)|y| \iff k^2 x^2 < (k + 1)^2 y^2$.
But $0 \leq kx^2 < (k + 1) y^2$ and $k \geq 0$, so $k^2 x^2 < (k + 1)^2 y^2$ follows easily.

\end{enumerate}



\subsection*{Extra notes}

The original problem only asks to prove that there exists a nice pair $(x, y)$ with $x \leq \sqrt{a}$ iff there exists a nice pair $(x, y)$ with $x > \sqrt{a}$.
Both clearly yields that a nice pair exists, and the existence of a nice pair implies one of the statements.
The above version shows a stronger result on one direction: if a nice pair exists, then one exists with arbitrarily large $x$-value.




\end{document}
