\documentclass{article}

\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{hyperref}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\newcommand{\N}{\mathbb{N}}

\newtheorem{lemma}{Lemma}

\title{IMO 2007 N6 (P5)}
\author{}
\date{}

\begin{document}

\maketitle



\subsection*{Problem}

Fix $n > 1$, and let $a$ and $b$ be positive integers such that $nab - 1 \mid (na^2 - 1)^2$.
Prove that $a = b$.



\subsection*{Solution}

Official solution: \url{https://www.imo-official.org/problems/IMO2007SL.pdf}

The official solution considers the case $n = 4$, but as in its comment section, any $n > 1$ works.
The following solution is essentially the official solution, presented in a slightly different way.

Fix $n > 1$, and call a pair $(a, b)$ of positive integers \emph{bad} if $a < b$ and $nab - 1 \mid (na^2 - 1)^2$.
The main results that we need for the descent argument are as follows.

\begin{lemma}\label{2007n6-1}
Let $a$ and be be positive integers such that $nab - 1 \mid (na^2 - 1)^2$.
Then $nab - 1 \mid (nb^2 - 1)^2$.
\end{lemma}
\begin{proof}
Indeed, $nab - 1$ divides $(na^2 - 1)^2 (nb^2)^2 = ((nab)^2 - nb^2)^2 = (nb^2 - (nab)^2)^2$, and $(nab)^2 \equiv 1 \pmod{nab - 1}$.
\end{proof}

\begin{lemma}\label{2007n6-2}
Fix some $a > 0$, and suppose that there exists $b > 0$ such that $(a, b)$ is a bad pair.
Then there exists an integer $c > 0$ such that $(c, a)$ is a bad pair.
\end{lemma}
\begin{proof}
Write $(na^2 - 1)^2 = (nab - 1) r$ for some integer $r$.
Working modulo $na$ gives $r \equiv -1 \pmod{na}$, so we can write $r = nac - 1$ for some integer $c$.
The bound $0 < a < b$ implies that $0 < c < a$, so it remains to show that $nac - 1 \mid (nc^2 - 1)^2$.
Indeed, we have $(na^2 - 1)^2 = (nab - 1)(nac - 1)$, so $nac - 1 \mid (na^2 - 1)^2$.
Then we are done using Lemma~\ref{2007n6-1}.
\end{proof}

If $nab - 1 \mid (na^2 - 1)^2$ holds for some positive integers $b < a$, then Lemma~\ref{2007n6-1} yields that $(b, a)$ is a bad pair.
Thus, to show that $nab - 1 \mid (na^2 - 1)^2$ implies $a = b$, it suffices to show that there exists no bad pairs.

Suppose for the sake of contradiction that bad pairs exist.
Let $a$ be the minimal positive integer such that $(a, b)$ is bad for some $b > a$.
By Lemma~\ref{2007n6-2}, there exists $0 < c < a$ such that $(c, a)$ is a bad pair.
The properties of such $c$ contradicts the minimality of $a$.
Thus, no bad pairs exist.



\subsection*{Implementation details}

We will use the \texttt{nat} datatype instead of an \texttt{int} datatype.
Aside, instead of implementing Lemma~\ref{2007n6-2} as a concrete lemma, we implement the descent argument as a lemma.
The lemma can be stated formally as follows.

\begin{lemma}\label{2007n6-3}
Let $P_0, P_1, P_2, \ldots$ be propositions.
Suppose that, for every $k \in \N$, $P_k$ implies $P_m$ for some $m \in \N$ such that $m < k$.
Then $P_k$ does not holds for any $k \in \N$.
\end{lemma}
\begin{proof}
Suppose not, and pick the minimal $k_0 \in \N$ such that $P_{k_0}$ holds.
Then $P_m$ holds for some $m \in \N$ with $m < k_0$.
This contradicts the minimality of $k_0$.
\end{proof}



\end{document}
