\documentclass{article}

\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{hyperref}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}

\title{IMO 2013 N2 (P1)}
\author{}
\date{}

\begin{document}

\maketitle



\subsection*{Problem}

Prove that, for any $k, n \in \N$, $n > 0$, there exists positive integers $m_1, m_2, \ldots, m_k$ such that
\[ 1 + \frac{2^k - 1}{n} = \prod_{j = 1}^k \left(1 + \frac{1}{m_j}\right). \tag{*}\label{2013n2-eq0} \]



\subsection*{Solution}

Official solution: \url{https://www.imo-official.org/problems/IMO2013SL.pdf}

We present Solution 1 of the official solution, with a slight modification.
We just induct on $k$.

Obviously, we are done if $k = 0$.
Now replace $k$ with $k + 1$, and consider two cases.

First, suppose that $n = 2t$ for some $t > 0$.
This time, we see that
\[ 1 + \frac{2^{k + 1} - 1}{2t} = \left(1 + \frac{1}{2(t + 2^k - 1)}\right) \left(1 + \frac{2^k - 1}{t}\right). \]
We are done by applying the induction hypothesis to the second factor.

Second, suppose that $n = 2t + 1$ for some $t \geq 0$.
Then we just notice that
\[ 1 + \frac{2^{k + 1} - 1}{2t + 1} = \left(1 + \frac{1}{2t + 1}\right) \left(1 + \frac{2^k - 1}{t + 1}\right). \]
Again, just apply the induction hypothesis to the second factor.



\subsection*{Implementation details}

The solution stored in \texttt{mathlib} uses \texttt{finset}.
Here, we will use \texttt{multiset} instead.

We will also use some definition hiding.
Given $k \geq 0$ and $c \in \N$, we say that $c$ is $k$-good if for any integer $n > 0$, there exists positive integers $m_1, m_2, \ldots, m_k$ such that
\[ 1 + \frac{c}{n} = \prod_{j = 1}^k \left(1 + \frac{1}{m_j}\right). \]
The original problem's statement asks to prove that $2^k - 1$ is $k$-good for any $k \geq 0$.
Solution 1 of the official solution, as presented above, goes more generally by proving that if $c$ is $k$-good, then $2c + 1$ is $(k + 1)$-good.
Another more obvious property is that $c$ being $k$-good implies that $c + 1$ is $(k + 1)$-good.



\end{document}
